Methods for Analyzing Models of Rod-Shaped Bacteria

### 2025-07-11 paper/main.tex
completely restructure paper; focus on methods; use estimation results only as an example

\begin{itemize}
    \item This publication contains 5 things:
    \begin{enumerate}
        \item Initial (incomplete) model of rod-shaped bacteria
        \item Methods of how to obtain data from microscopic images and perform parameter
            estimations; $\rightarrow$ also possibly across multiple generations
        \item Showcase the usefulness of the methods and the model by applying it to parameter
            estimation via microscopic images and multilayer formation
        \item Generate 3D Images (masks)
        \item A python package which can do all of that and can be reused and extended
    \end{enumerate}
    \item Make clear that this is mainly a proof of the methods
    \item For complete model, more work has to be done
    \begin{itemize}
        \item More aspects (see Review; put in discussion)
        \item Note that we only care about \textit{E.Coli} and \textit{B.Subtilits}! This is
            important.
    \end{itemize}
\end{itemize}

\begin{figure}
    \centering
    \tikzstyle{rect} = [
        rectangle,
        % rounded corners,
        minimum width=4cm,
        minimum height=1cm,
        text centered,
        draw=black,
        fill=color4!70,
    ]
    \tikzstyle{frect} = [rect, fill=color3!40]
    \tikzstyle{arrow} = [thick,->,>=stealth]
    \tikzstyle{darrow} = [arrow, dashed]
    \begin{tikzpicture}
        \node (model) [rect, align=center] at (0,0) {Mathematical \&\\ Computational Model};
        \node (extraction) [rect] at (0, -2) {Data Extraction};
        \node (params) [rect] at (6, -2) {Parameter Estimation};
        \node (visualization) [rect] at (6, 0) {Visualization};

@@

        \draw[arrow] (visualization) -- (realimages);
        \draw[arrow] (realimages) -- (segmentation);
        \draw[arrow] (extraction) -- (analysis);
        \draw[arrow] (params) -- (realimages);
    \end{tikzpicture}
    \caption{
        The project can be divided into 4 subparts (yellow) that interact with and depend on each
        other.
        Additional blocks (blue-gray) indicate possible future extensions of this project.
        Arrows with solid lines indicate dependencies between the subparts while dashed arrows carry
        special meaning, which is specifically annotated.
        The mathematical models provides the theoretical framework and is thus the parent of all
        following work.
        "Data Extraction" allows us to read variable and parameter values directly from microscopic
        images.
        The obtained data can then be used to initialize a numerical simulations and thus estimate
        parameters.
        On the other hand, results produced by our model can be visualized as 3D renders and
        projected to cell-masks.
        These algorithms can be exploited to construct cost-functions which can be used in the
        parameter estimation process.
        In future work, this visualization scheme could be extended to generate realistic
        microscopic images with cell-masks and thus train cell-segmentation and cell-tracking
        algorithms.
    }
    \label{fig:flowchart-project-structure}
\end{figure}

@@

%###################################################################################################
\section{A mechanical Model of Rod-Shaped Bacteria}
\label{section:mechanical-model-rod-shaped-bacteria}

@@

\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{docs/source/_static/mechanics.png}
    \caption{
        Example of an elonged bacterial rod represented by a collection of vertices which are
        connected by springs.
        The angles $\alpha_i$ have been exaggerated such that they can be identified more easily.
        A bacterial agent which is not interacting with any external objects will tend to the
        equilibrium state of a straight rod with $0$ curvature.
    }
    \label{fig:mechanics-bacterium}
\end{figure}

% --------------------------------------------------------------------------------------------------
\subsection{Interactions}

@@

% --------------------------------------------------------------------------------------------------
\subsection{Cycle}

@@

% --------------------------------------------------------------------------------------------------
\subsection{External Forces}
TODO
\begin{itemize}
    \item Friction force against surface. perpendicular to motion; proportional to velocity.
    \item gel pressure coming from experiment (simply downwards force $F_z$)
\end{itemize}

@@

%###################################################################################################
\section{Generating Masks and Images}
\label{section:generating-masks-and-images}
\begin{itemize}
    \item use pyvista (VTK) for Visualization
    \item always do 3D render. This way, we can have overalp between agents and produce more
        realistic masks.
    \item Assign unique color values for idividual agents.
        We have a 1:1 mapping between agent identifier and color! (for each simulation different
        though)
    \item use parallel projection; this should be most close to what microscopes measure due to the
        relatively large distance between object and lens.
    \item use some noise to smear out the rods; this is already similar to microscopic images; use
        more like this and deal with real-world defects to make it more realistic (future, not now)
    \item Figure~\ref{fig:progression-image-generation} shows steps in generating image.
    \item Reflectivity, placement of light etc. can already introduce some heterogeneity in the
        visual of the rod.
    \item Plan to implement noise on the mesh itself; make it have rough surface (like real rods as
        seen under electron microscopy)
\end{itemize}

%###################################################################################################
\section{Parameter Estimation}

In order to effectively estimate the parameters of our model, we need to set up a cost function
which can be minimized.
In doing so, we hope to identify all parameters, given relevant experimental data.
We use microscopic images and construct a novel algorithm that can extract the position of the agent
from a provided cell-mask.
These cell-masks are easily obtainable by either using existing segmentation tools or manually
annottating images.
Following this, we can define a cost-function and an optimization scheme to minimize said function.

% --------------------------------------------------------------------------------------------------
\subsection{Extract Positions from Masks}
\label{section:fitting-and-comparing-masks}

% --------------------------------------------------------------------------------------------------
\subsection{Benchmarking the Extraction Algorithm}
\label{subsection:benchmarking-extraction-algorithm}

% --------------------------------------------------------------------------------------------------
\subsection{Estimating Parameters}
\label{section:parameter-estimation}

We will now focus on using the presented methods for parameter estimation via microscopic images.
Our initial state is given by TODO~\cite{}.% TODO
We use our extraction algorithm to obtain the initial positions of our cells.
From there, we supply the model with parameters such that we can make a prediction for the final
positions of the bacteria.
We chose a time-interval which does not contain any division events.
By doing so, we can utilize the simpler and more performant position comparison scheme for our cost
function.
Another important detail is the individuality of the agents.
In our case, this means that each agent might have different parameters associated to it.
For this dataset, we assumed that all interaction parameters which specify the interaction
potential are identical across all agents but growth rates and the physical radii of the cells are
individual and we also optimize them.
We chose to optimize for 3 interaction parameters of the Mie potential $n,m,V_0$ and two parameters
which were individual for each cell, namely the radii $r_i$ and growth rates $\mu_i$.
We also included the damping constant $\lambda$ and removed any stochastic motion contributions
($D=0$).
This leads to a total of $4+2\times 6=16$ parameters.
To estimate these parameters we used a latin-hypercube sweep across the parameter space and
performed a local minimization with the Nelder-Mead method (TODO citations).% CITATIONS

@@

>>>>

\begin{itemize}
    \item This publication contains 5 things:
    \begin{enumerate}
        \item Methods of how to obtain data from microscopic images and perform parameter
            estimations; $\rightarrow$ also possibly across multiple generations
        \item Showcase the usefulness of the methods and the model by applying it to parameter
            estimation via microscopic images and multilayer formation
        \item Initial (incomplete) model of rod-shaped bacteria
        \item Generate 3D Images (masks)
        \item A python package which can do all of that and can be reused and extended
    \end{enumerate}
    \item Overall structure
    \begin{enumerate}
        \item users bring model + data
        \item this paper provides methods to do the parameter estimation + visualization
    \end{enumerate}
    \item For complete model, more work has to be done
    \begin{itemize}
        \item More aspects (see Review; put in discussion)
        \item Note that we only care about \textit{E.Coli} and \textit{B.Subtilits}! This is
            important.
    \end{itemize}
\end{itemize}

\begin{figure}[H]
    \centering
    \tikzstyle{rect} = [
        rectangle,
        % rounded corners,
        minimum width=4cm,
        minimum height=1cm,
        text centered,
        draw=black,
        fill=color4!70,
    ]
    \tikzstyle{brect} = [rect, fill=color5!50]
    \tikzstyle{frect} = [rect, fill=color3!40]
    \tikzstyle{arrow} = [thick,->,>=stealth]
    \tikzstyle{darrow} = [arrow, dashed]
    \begin{tikzpicture}
        \node (model) [brect, align=center] at (0,0) {Mathematical \&\\ Computational Model};
        \node (extraction) [rect] at (0, -2) {Data Extraction};
        \node (params) [brect] at (6, -2) {Parameter Estimation};
        \node (visualization) [rect] at (6, 0) {Visualization};

@@

        \draw[arrow] (visualization) -- (realimages);
        \draw[arrow] (realimages) -- (segmentation);
        \draw[arrow] (extraction) -- (analysis);
        \draw[arrow] (params) -- (realimages);
    \end{tikzpicture}
    \caption{
        The project can be divided into 4 subparts (orange + yellow) that interact with and depend
        on each other.
        The orange blocks require input by the user by specifying the model and supplying input
        data.
        Additional blocks (blue-gray) indicate possible future extensions of this project.
        Arrows with solid lines indicate dependencies between the subparts while dashed arrows carry
        special meaning, which is specifically annotated.
        The mathematical models provides the theoretical framework and is thus the parent of all
        following work.
        "Data Extraction" allows us to read variable and parameter values directly from microscopic
        images.
        The obtained data can then be used to initialize a numerical simulations and thus estimate
        parameters.
        On the other hand, results produced by our model can be visualized as 3D renders and
        projected to cell-masks.
        These algorithms can be exploited to construct cost-functions which can be used in the
        parameter estimation process.
        In future work, this visualization scheme could be extended to generate realistic
        microscopic images with cell-masks and thus train cell-segmentation and cell-tracking
        algorithms.
    }
    \label{fig:flowchart-project-structure}
\end{figure}

@@

%###################################################################################################
\section{Modeling Prerequesites}
\begin{enumerate}
    \item represent bacteria as polygon with vertices $\vec{v}_i$
    \item Assumes that these vertices are not fundamental but rather a discretization of the actual
        problem
\end{enumerate}

\begin{figure}[H]
    \centering
    \includegraphics[width=0.5\textwidth]{docs/source/_static/mechanics.png}
    \caption{
        Example of an elonged bacterial rod represented by a collection of vertices which are
        connected by springs.
        The angles $\alpha_i$ have been exaggerated such that they can be identified more easily.
        A bacterial agent which is not interacting with any external objects will tend to the
        equilibrium state of a straight rod with $0$ curvature.
    }
    \label{fig:mechanics-bacterium}
\end{figure}

%###################################################################################################
\section{Data Extraction}
\label{section:fitting-and-comparing-masks}

% --------------------------------------------------------------------------------------------------
\paragraph{Benchmarking the Extraction Algorithm}
\label{subsection:benchmarking-extraction-algorithm}

%###################################################################################################
\section{Visualization}
\label{section:generating-masks-and-images}
\begin{itemize}
    \item use pyvista (VTK) for Visualization
    \item always do 3D render. This way, we can have overalp between agents and produce more
        realistic masks.
    \item Assign unique color values for idividual agents.
        We have a 1:1 mapping between agent identifier and color! (for each simulation different
        though)
    \item use parallel projection; this should be most close to what microscopes measure due to the
        relatively large distance between object and lens.
    \item use some noise to smear out the rods; this is already similar to microscopic images; use
        more like this and deal with real-world defects to make it more realistic (future, not now)
    \item Figure~\ref{fig:progression-image-generation} shows steps in generating image.
    \item Reflectivity, placement of light etc. can already introduce some heterogeneity in the
        visual of the rod.
    \item Plan to implement noise on the mesh itself; make it have rough surface (like real rods as
        seen under electron microscopy)
\end{itemize}

%###################################################################################################
\section{Parameter Estimation}
In order to effectively estimate the parameters of our model, we need to set up a cost function
which can be minimized.
In doing so, we hope to identify all parameters, given relevant experimental data.
We use microscopic images and construct a novel algorithm that can extract the position of the agent
from a provided cell-mask.
These cell-masks are easily obtainable by either using existing segmentation tools or manually
annottating images.
Following this, we can define a cost-function and an optimization scheme to minimize said function.

\section{Applications}
% --------------------------------------------------------------------------------------------------
\subsection{A Mechanical Model of Rod-Shaped Bacteria}
\label{subsubsection:mechanical-model-rod-shaped-bacteria}

%###################################################################################################

@@

% --------------------------------------------------------------------------------------------------
\subsection{Interactions}
\label{subsection:mechanical-model-interactions}

@@

% --------------------------------------------------------------------------------------------------
\subsection{Cycle}
\label{subsection:mechanical-model-cycle}

@@

% --------------------------------------------------------------------------------------------------
\subsection{External Forces}
\label{subsection:mechanical-model-external-forces}
TODO
\begin{itemize}
    \item Friction force against surface. perpendicular to motion; proportional to velocity.
    \item gel pressure coming from experiment (simply downwards force $F_z$)
\end{itemize}

@@

% --------------------------------------------------------------------------------------------------
\subsection{Estimating Mechanical Parameters of Growing \textit{E.Coli.} on Agar}
We will now focus on using the presented methods for parameter estimation via microscopic images.
Our initial state is given by TODO~\cite{}.% TODO
We use our extraction algorithm to obtain the initial positions of our cells.
From there, we supply the model with parameters such that we can make a prediction for the final
positions of the bacteria.
We chose a time-interval which does not contain any division events.
By doing so, we can utilize the simpler and more performant position comparison scheme for our cost
function.
Another important detail is the individuality of the agents.
In our case, this means that each agent might have different parameters associated to it.
For this dataset, we assumed that all interaction parameters which specify the interaction
potential are identical across all agents but growth rates and the physical radii of the cells are
individual and we also optimize them.
We chose to optimize for 3 interaction parameters of the Mie potential $n,m,V_0$ and two parameters
which were individual for each cell, namely the radii $r_i$ and growth rates $\mu_i$.
We also included the damping constant $\lambda$ and removed any stochastic motion contributions
($D=0$).
This leads to a total of $4+2\times 6=16$ parameters.
To estimate these parameters we used a latin-hypercube sweep across the parameter space and
performed a local minimization with the Nelder-Mead method (TODO citations).% CITATIONS

@@

% --------------------------------------------------------------------------------------------------
\subsection{Data Generation}
